Calendar-based graphics for visualizing people's daily schedules

### 2017-08-03 main.Rmd
many tweaks from Rob's feedback

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

We develop a method for organizing and visualizing temporal data, collected on sub-daily interval, into a calendar layout. The purpose of the calendar-based visualization is to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area (Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily rhythms, to assist the city administration and local businesses with event planning and operational management. A routine examination of the data would be constructing a conventional time series plot to catch a glimpse of temporal patterns. The faceted plots in Figure \ref{fig:time-series-plot}, give an overall picture of the foot traffic at 3 different sensors over 2016. Further faceting by day of the week (Figure \ref{fig:facet-time}) provides a better glimpse at the daily and sub-daily, like commuter, patterns. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

1. Variations primarily source from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurring events).
2. Since the data are often collected at sub-daily frequencies, they typically involve a large number of observations.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need for comparing and contrasting between locations.

@@

(ref:ped-map-cap) Map of the Melbourne city with purple dots indicating sensor locations.

@@

(ref:time-series-plot-cap) Time series plot about the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these 3 sensors. There is an eye-catching spike which occurred at State Library, caused by the annual event--White Night--on 20th of February.

@@

(ref:facet-time-cap) Hourly pedestrian counts faceted by sensors and days of the week using lines. It features at least two types of seasons--time of day and day of week--across all the sensors. The temporal patterns are subject to the sensor locations too.

@@

In constrast, calendar-based graphics unpacks the temporal variable, at different resolutions, to digest multiple seasonalities, and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasons like school vacations; all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout, (3) to enable an overlaying plots consisting of multiple time series. The proposed algorithm has been implemented in the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is an implementation of the grammar of graphics [@wilkinson2006grammar; @wickham2010layered].

@@

The algorithm for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar, requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month can be captured with 35 (5 $\times$ 7) cells, where the top left is Monday of week 1, and the bottom right is Sunday of week 5 by default. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday-Sunday, which is determined by the year of the calendar. Months are of different length days, ranging from 28 to 31, and each month could extend over six weeks but the convention in these months is to wrap the last few days up to the top row of the block. The notation for creating these cells is as follows:

@@

There are several options provided for the \code{frame_calendar} function to initialize and adjust the display of a calender plot:

@@

The algorithm described in Section \ref{sec:algorithm} is for the most common calendar layout, the "monthly" calendar, but it can be simplified to accommodate the other two types of calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as the "weekly" and "daily" calendar, respectively. The layout to be used is controlled by the \code{calendar} argument. The weekly calendar puts more emphasis on days of a week over days of a year, whereas the daily calender does the opposite. The monthly calender can be considered as the advanced variant of both weekly and daily. Temporal patterns motivate which variant should be employed. The weekly calendar is appropriate if most of the variation can be characterised by days of the week. On the other hand, the daily calendar should be used when there is a yearly effect but not a weekly effect in the data. When both effects are present, the monthly calendar would be a better choice.

@@

When \code{polar = TRUE}, the polar transformation is carried out on the data. The computation is similar to the one described in @Wickham2012glyph. Figure \ref{fig:flinders-polar} shows star plots embedded in the monthly calendar layout, which is equivalent to Figure \ref{fig:flinders-2016} placed in polar coordinates.

@@

The \code{free_wday} suggests that the scales are common for a given week day across all weeks, but different for each week days within a week. In other words, the same value within a given Monday, for example, corresponds to the same position across all Mondays but the different position in other week days. Similarly, the \code{free_mday} suggests free scaling for any day within a given month. The scaling choices combined with the calendar layouts offers a number of varieties to construct the plot.

@@

(ref:flinders-free-cap) Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive, however it is more difficult to compare the size of peaks between days.

@@

The comparison of one sensor to another adds additional insights to the dataset, which is often done with an overlaying plot such as Figure \ref{fig:overlay}. For instance, the dominant patterns that occurred at Flinders Street and Flagstaff train stations are driven by the commuters on work days; however, Flinders Street Station has higher pedestrian counts during the weekends and public holidays. This suggests that Flagstaff Station limits its functionality on non-work days, but various activities take place around the Flinders Street Station other than commuting. Figure \ref{fig:overlay} also shows that State Library follows a similar temporal trend as Flinders Street Station on non-work days. The nighttime events, such as White Night and New Year's Eve, have barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to Flinders Street Station and State Library.

@@

The \code{frame_calendar} function does not constrain itself to mapping only the temporal variable to the \code{x} argument. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organized in the daily calendar layout. It provides a visual tool for identifying repeating patterns. Figure \ref{fig:scatterplot} indicates two separate paths in the work-day glyphs, suggesting that an hour of a work day with many pedestrians is likely to be followed in the next hour with either greater or fewer numbers of pedestrians. However, this pattern is absent on non-work days. There is a bimodality on work days while there is unimodality on the rest of the days, which is also supported by Figure \ref{fig:flinders-2016}.

@@

The newly computed coordinates can work with more complicated plots, such as the boxplot. Figure \ref{fig:boxplot} uses the loess smooth line superimposed on side-by-side boxplots. It shows the distribution of hourly counts across all 43 sensors at a given time point during December. In general, bimodality features on work days whereas unimodality features on the rest of the days. The last week of December is the holiday season: people are off work on the day before Christmas, go shopping on the Boxing day, and stay out for the fireworks on New Year's Eve.

@@

These calendar layouts, especially the monthly calendar, become widely used in society. We offer languages other than English for text labelling. Figure \ref{fig:chn} shows the same plot as Figure \ref{fig:boxplot} labelled using simplified Chinese characters.

@@

This sort of layout may be useful for studying consumer trends, or human behavior, like the pedestrian patterns. It may not work so well for physical patterns like temperature, which are not typically affected by human activity. Additionally, it is perceptually difficult to discover an overall trend of less significance, since there is the lack of a common baseline due to the rows and columns of a display matrix.

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} uses linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

We develop a method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout. The purpose of the calendar-based visualization is to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area (Figure \ref{fig:ped-map}). The dataset can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. A routine examination of the data would involve constructing conventional time series plots to catch a glimpse of temporal patterns. The faceted plots in Figure \ref{fig:time-series-plot}, give an overall picture of the foot traffic at 3 different sensors over 2016. Further faceting by day of the week (Figure \ref{fig:facet-time}) provides a better glimpse of the daily and sub-daily pedestrian patterns. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

1. Variations exist on multiple time scales including time of day, day of week, and day of year (such as public holiday and recurring events).
2. Since the data are often collected at sub-daily frequencies, they typically involve a large number of observations.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need for comparing and contrasting between locations.

@@

(ref:ped-map-cap) Map of the Melbourne city area with purple dots indicating sensor locations.

@@

(ref:time-series-plot-cap) Time series plot showing the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these 3 sensors. There is an eye-catching spike which occurred at the State Library, caused by the annual White Night event on 20th of February.

@@

(ref:facet-time-cap) Hourly pedestrian counts faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors.

@@

In contrast, calendar-based graphics unpacks the temporal variable, at different resolutions, to digest multiple seasonalities, and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm to go beyond the calendar-based heatmap. The approach is developed with three conditions in mind: (1) to display time-of-day variation in addition to longer temporal components such as day-of-week and day-of-year; (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout; (3) to enable overlaying plots consisting of multiple time series. The proposed algorithm has been implemented in the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is an implementation of the grammar of graphics [@wilkinson2006grammar; @wickham2010layered].

@@

The algorithm for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month can be captured with 35 (5 $\times$ 7) cells, where the top left is Monday of week 1, and the bottom right is Sunday of week 5 by default. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is determined by the year of the calendar. Months are of different lengths, ranging from 28 to 31 days, and each month could extend over six weeks but the convention in these months is to wrap the last few days up to the top row of the block. The notation for creating these cells is as follows:

@@

There are several options provided for the \code{frame_calendar} function to initialize and adjust the display of a calendar plot:

@@

The algorithm described in Section \ref{sec:algorithm} is for the most common calendar layout, the "monthly" calendar, but it can be simplified to accommodate two other types calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as the "weekly" and "daily" calendar, respectively. The layout to be used is controlled by the \code{calendar} argument. The weekly calendar puts more emphasis on days of a week over days of a year, whereas the daily calendar does the opposite. The monthly calendar can be considered as the advanced variant of both weekly and daily. Temporal patterns motivate which variant should be employed. The weekly calendar is appropriate if most of the variation can be characterised by days of the week. On the other hand, the daily calendar should be used when there is a yearly effect but not a weekly effect in the data (for example weather data). When both effects are present, the monthly calendar would be a better choice.

@@

When \code{polar = TRUE}, a polar transformation is carried out on the data. The computation is similar to the one described in @Wickham2012glyph. Figure \ref{fig:flinders-polar} shows star plots embedded in the monthly calendar layout, which is equivalent to Figure \ref{fig:flinders-2016} placed in polar coordinates.

@@

The \code{free_wday} suggests that the scales are common for a given week day across all weeks, but different for each week day within a week. In other words, the same value within a given Monday, for example, corresponds to the same position across all Mondays but a different position in other week days. Similarly, the \code{free_mday} suggests free scaling for any day within a given month. The scaling choices combined with the calendar layouts offers a number of varieties to construct the plot.

@@

(ref:flinders-free-cap) Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive, however it is impossible to compare the size of peaks between days.

@@

The comparison of one sensor to another adds additional insights to the dataset, which is often done with an overlaying plot such as Figure \ref{fig:overlay}. For instance, the dominant patterns that occurred at Flinders Street and Flagstaff train stations are driven by the commuters on work days; however, Flinders Street Station has higher pedestrian counts during the weekends and public holidays. This suggests that Flagstaff Station limits its functionality on non-work days, but various activities take place around the Flinders Street Station other than commuting. Figure \ref{fig:overlay} also shows that the State Library follows a similar temporal trend as Flinders Street Station on non-work days. The nighttime events, such as White Night and New Year's Eve, have barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to Flinders Street Station and the State Library.

@@

The \code{frame_calendar} function is not constrained to mapping only the temporal variable to the \code{x} argument. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organized in the daily calendar layout. It provides a visual tool for identifying repeating patterns. Figure \ref{fig:scatterplot} indicates two separate paths in the work-day glyphs, suggesting that an hour of a work day with many pedestrians is likely to be followed in the next hour with either greater or fewer numbers of pedestrians. However, this pattern is absent on non-work days. There is a bimodality on work days while there is unimodality on the rest of the days, which is also supported by Figure \ref{fig:flinders-2016}.

@@

The newly computed coordinates can also work with more complicated plots, such as boxplots. Figure \ref{fig:boxplot} uses a loess smooth line superimposed on side-by-side boxplots. It shows the distribution of hourly counts across all 43 sensors during December. The last week of December is the holiday season: people are off work on the day before Christmas, go shopping on the Boxing day, and stay out for the fireworks on New Year's Eve.

@@

We also offer languages other than English for text labelling. Figure \ref{fig:chn} shows the same plot as Figure \ref{fig:boxplot} labelled using simplified Chinese characters.

@@

This sort of layout may be useful for studying consumer trends, or human behavior, such as pedestrian patterns. It may not work so well for physical patterns like temperature, which are not typically affected by human activity. Additionally, it is perceptually difficult to discover an overall trend, since there is no common baseline due to the rows and columns of a display matrix.

### 2017-08-03 main.tex
many tweaks from Rob's feedback

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly intervals by sensors scattered around the city.
Faceting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format options and variations. It is implemented in the R
package \pkg{sugrrants}.
}

@@

We develop a method for organizing and visualizing temporal data,
collected on sub-daily interval, into a calendar layout. The purpose of
the calendar-based visualization is to provide insights into people's
daily schedules relative to events such as work days, weekends,
holidays, and special events. This work was originally motivated by
studying foot traffic in the city of Melbourne \citep{ped}. There have
been 43 sensors installed that count pedestrians every hour across the
downtown area (Figure \ref{fig:ped-map}). The dataset can shed light
into understanding people's daily rhythms, to assist the city
administration and local businesses with event planning and operational
management. A routine examination of the data would be constructing a
conventional time series plot to catch a glimpse of temporal patterns.
The faceted plots in Figure \ref{fig:time-series-plot}, give an overall
picture of the foot traffic at 3 different sensors over 2016. Further
faceting by day of the week (Figure \ref{fig:facet-time}) provides a
better glimpse at the daily and sub-daily, like commuter, patterns. The
sensor data, like many temporal datasets on human behaviors, lends
itself to a number of exploratory data visualization challenges:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily source from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurring events).
\item
  Since the data are often collected at sub-daily frequencies, they
  typically involve a large number of observations.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need for comparing and contrasting
  between locations.
\end{enumerate}

@@

\caption[Map of the Melbourne city with purple dots indicating
sensor locations.]{Map of the Melbourne city with purple dots indicating
sensor locations.}\label{fig:ped-map}
\end{figure}
\end{CodeChunk}

@@

\caption[Time series plot about the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Coloured by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at State Library, caused
by the annual event--White Night--on 20th of February.]{Time series plot about the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Coloured by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at State Library, caused
by the annual event--White Night--on 20th of February.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

\caption[Hourly pedestrian counts faceted by sensors and
days of the week using lines. It features at least two types of
seasons--time of day and day of week--across all the sensors. The
temporal patterns are subject to the sensor locations too.]{Hourly pedestrian counts faceted by sensors and
days of the week using lines. It features at least two types of
seasons--time of day and day of week--across all the sensors. The
temporal patterns are subject to the sensor locations too.}\label{fig:facet-time}
\end{figure}
\end{CodeChunk}

@@

In constrast, calendar-based graphics unpacks the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasons like school vacations; all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in two R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are limited to
colour-encoding graphics and are unable to use time scales smaller than
day. Time of day, which serves as one of the most important aspects in
explaining substantial variations arising from pedestrian sensor data,
will be neglected through daily aggregation. Additionally, if simply
using colored blocks rather than curves, it may become perceptually
difficult to estimate the shape positions and changes, although using
curves comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to
incorporate line graphs and other types of glyphs into the graphical
toolkit for the calendar layout, (3) to enable an overlaying plots
consisting of multiple time series. The proposed algorithm has been
implemented in the \code{frame_calendar} function in the \pkg{sugrrants}
package \citep{R-sugrrants} using \proglang{R} \citep{R-base}.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is an
implementation of the grammar of graphics
\citep{wilkinson2006grammar, wickham2010layered}.

@@

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar, requires cells
for days, embedded in blocks corresponding to months, organized into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5 by default. These cells provide a micro canvas on
which to plot the data. The first day of the month could be any of
Monday-Sunday, which is determined by the year of the calendar. Months
are of different length days, ranging from 28 to 31, and each month
could extend over six weeks but the convention in these months is to
wrap the last few days up to the top row of the block. The notation for
creating these cells is as follows:

@@

The algorithm described in Section \ref{sec:algorithm} is for the most
common calendar layout, the ``monthly'' calendar, but it can be
simplified to accommodate the other two types of calendar formats. One
is comprised of days of a week in columns and weeks of a year in rows,
and the other is days of a month in columns and months of a year in
rows, which we refer to as the ``weekly'' and ``daily'' calendar,
respectively. The layout to be used is controlled by the \code{calendar}
argument. The weekly calendar puts more emphasis on days of a week over
days of a year, whereas the daily calender does the opposite. The
monthly calender can be considered as the advanced variant of both
weekly and daily. Temporal patterns motivate which variant should be
employed. The weekly calendar is appropriate if most of the variation
can be characterised by days of the week. On the other hand, the daily
calendar should be used when there is a yearly effect but not a weekly
effect in the data. When both effects are present, the monthly calendar
would be a better choice.

@@

When \code{polar = TRUE}, the polar transformation is carried out on the
data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} shows star
plots embedded in the monthly calendar layout, which is equivalent to
Figure \ref{fig:flinders-2016} placed in polar coordinates.

@@

\caption[Line glyphs on the calendar format showing
hourly foot traffic at Flinders Street Station, scaled over all the
days. The individual shape on a single day becomes more distinctive,
however it is more difficult to compare the size of peaks between days.]{Line glyphs on the calendar format showing
hourly foot traffic at Flinders Street Station, scaled over all the
days. The individual shape on a single day becomes more distinctive,
however it is more difficult to compare the size of peaks between days.}\label{fig:flinders-free}
\end{figure}
\end{CodeChunk}

>>>>

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} uses linear
algebra on the date variable to create the layout. It utilizes the
grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior in Melbourne, Australia,
based on counts which are captured at hourly intervals by sensors
scattered around the city. Faceting by the usual features such as day
and month, was insufficient to examine the behavior. Making displays on
a monthly calendar format helps to understand pedestrian patterns
relative to events such as work days, weekends, holidays, and special
events. The layout algorithm has several format options and variations.
It is implemented in the R package \pkg{sugrrants}.
}

@@

We develop a method for organizing and visualizing temporal data,
collected at sub-daily intervals, into a calendar layout. The purpose of
the calendar-based visualization is to provide insights into people's
daily schedules relative to events such as work days, weekends,
holidays, and special events. This work was originally motivated by
studying foot traffic in the city of Melbourne, Australia \citep{ped}.
There have been 43 sensors installed that count pedestrians every hour
across the downtown area (Figure \ref{fig:ped-map}). The dataset can
shed light on people's daily rhythms, and assist the city administration
and local businesses with event planning and operational management. A
routine examination of the data would involve constructing conventional
time series plots to catch a glimpse of temporal patterns. The faceted
plots in Figure \ref{fig:time-series-plot}, give an overall picture of
the foot traffic at 3 different sensors over 2016. Further faceting by
day of the week (Figure \ref{fig:facet-time}) provides a better glimpse
of the daily and sub-daily pedestrian patterns. The sensor data, like
many temporal datasets on human behaviors, lends itself to a number of
exploratory data visualization challenges:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations exist on multiple time scales including time of day, day of
  week, and day of year (such as public holiday and recurring events).
\item
  Since the data are often collected at sub-daily frequencies, they
  typically involve a large number of observations.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need for comparing and contrasting
  between locations.
\end{enumerate}

@@

\caption[Map of the Melbourne city area with purple dots
indicating sensor locations.]{Map of the Melbourne city area with purple dots
indicating sensor locations.}\label{fig:ped-map}
\end{figure}
\end{CodeChunk}

@@

\caption[Time series plot showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Coloured by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.]{Time series plot showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Coloured by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

\caption[Hourly pedestrian counts faceted by sensors and
days of the week using lines. It features at least two types of
seasons---time of day and day of week---across all the sensors.]{Hourly pedestrian counts faceted by sensors and
days of the week using lines. It features at least two types of
seasons---time of day and day of week---across all the sensors.}\label{fig:facet-time}
\end{figure}
\end{CodeChunk}

@@

In contrast, calendar-based graphics unpacks the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasonal variation such as school vacations, all of which have
influence over the turn-outs in the office. The calendar-based heatmap
was implemented in two R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are limited to colour-encoding graphics and are unable to use
time scales smaller than a day. Time of day, which serves as one of the
most important aspects in explaining substantial variations arising from
pedestrian sensor data, will be neglected through daily aggregation.
Additionally, if simply using colored blocks rather than curves, it may
become perceptually difficult to estimate the shape positions and
changes, although using curves comes with the cost of more display
capacity \citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm to go beyond the calendar-based heatmap. The
approach is developed with three conditions in mind: (1) to display
time-of-day variation in addition to longer temporal components such as
day-of-week and day-of-year; (2) to incorporate line graphs and other
types of glyphs into the graphical toolkit for the calendar layout; (3)
to enable overlaying plots consisting of multiple time series. The
proposed algorithm has been implemented in the \code{frame_calendar}
function in the \pkg{sugrrants} package \citep{R-sugrrants} using
\proglang{R} \citep{R-base}.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is an
implementation of the grammar of graphics
\citep{wilkinson2006grammar, wickham2010layered}.

@@

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar requires cells
for days, embedded in blocks corresponding to months, organized into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5 by default. These cells provide a micro canvas on
which to plot the data. The first day of the month could be any of
Monday--Sunday, which is determined by the year of the calendar. Months
are of different lengths, ranging from 28 to 31 days, and each month
could extend over six weeks but the convention in these months is to
wrap the last few days up to the top row of the block. The notation for
creating these cells is as follows:

@@

The algorithm described in Section \ref{sec:algorithm} is for the most
common calendar layout, the ``monthly'' calendar, but it can be
simplified to accommodate two other types calendar formats. One is
comprised of days of a week in columns and weeks of a year in rows, and
the other is days of a month in columns and months of a year in rows,
which we refer to as the ``weekly'' and ``daily'' calendar,
respectively. The layout to be used is controlled by the \code{calendar}
argument. The weekly calendar puts more emphasis on days of a week over
days of a year, whereas the daily calendar does the opposite. The
monthly calendar can be considered as the advanced variant of both
weekly and daily. Temporal patterns motivate which variant should be
employed. The weekly calendar is appropriate if most of the variation
can be characterised by days of the week. On the other hand, the daily
calendar should be used when there is a yearly effect but not a weekly
effect in the data (for example weather data). When both effects are
present, the monthly calendar would be a better choice.

@@

When \code{polar = TRUE}, a polar transformation is carried out on the
data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} shows star
plots embedded in the monthly calendar layout, which is equivalent to
Figure \ref{fig:flinders-2016} placed in polar coordinates.

@@

\caption[Line glyphs on the calendar format showing
hourly foot traffic at Flinders Street Station, scaled over all the
days. The individual shape on a single day becomes more distinctive,
however it is impossible to compare the size of peaks between days.]{Line glyphs on the calendar format showing
hourly foot traffic at Flinders Street Station, scaled over all the
days. The individual shape on a single day becomes more distinctive,
however it is impossible to compare the size of peaks between days.}\label{fig:flinders-free}
\end{figure}
\end{CodeChunk}
